---
title: "Intro to R/R Studio"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = TRUE)
knitr::opts_knit$set(root.dir = '../')
```

# Installing R and Rstudio
R can be installed from [here](https://cran.r-project.org). Download the base
version
for your operating system.  
RStudio can be installed from 
[here](https://www.rstudio.com/products/rstudio/download/#download). Download
the
installer for your platform.  

**R must be installed before installing RStudio**  

# R code and R output  
Throughout this course you will be presented with R code and R output. 
R code will be in boxes like this:
```{r, eval=FALSE}
print('This is some R code.')
```

This can be copied and pasted as is into your R script or R console. 

R output will be preceded by double hash marks and look like this:
```{r, echo=F}
print('This is R output.')
```

Comments can be added to R code with a hash. Your code should be liberally 
commented. R will ignore anything in a comment. Comments can also be added to 
the end of code lines.  

```{r}
# This is a comment
x <- 5 # This assigns the number 5 to x.
```

## R is case sensitive  
Function names, object names, etc. are all case sensitive in R. This means that 
the function mean() $\neq$ Mean() $\neq$ MEAN() and data $\neq$ Data $\neq$ DATA.
```{r}
a <- 1
A <- 5
```

```{r}
a
A
identical(a, A)
```

## Quotation Marks
Character strings can be put in quotation marks as ' or ", but they must be 
closed the same way. If a string contains a quotation mark the opposite style 
should be used to open and close it. For instance, in strings with an 
apostrophe (single quotation), double quotation marks should be used to open 
and close that string. Otherwise it is user preference. 

```{r}
print('Hello')
print("Hello")

```

```{r}
print('This gives an error")
```


```{r}
print("Don't use single if there's an apostrophe")
print('and "vice versa"')
```


# File Paths  
A file path specifies the unique location of a file. For example the file path 
for this file is '/Users/caroline/Repositories/BIOS6605/Modules/Module_1.html'. 
R recognizes relative file paths. For example if the current working directory 
(see below) is '/Users/caroline/Repositories/BIOS6605/', R will allow you to 
access this file with the relative file path 'Modules/Module_1.html'.  

## Some Handy Shortcuts  
'~/' always represents a users home folder and '../' takes you back up one
folder level. '../' can be used in succession to go up multiple folder levels. 
For example this file's path can be represented as 
'~/Repositories/BIOS6605/Modules/Module_1.html' (where '/Users/caroline' is 
replaced with '~') '../' will take you to '/Users/caroline/Repositories/' (one 
level up from '/Users/caroline/Repositories/BIOS6605/') and '../../' will take you
to '/Users/caroline/', two levels up.

## Slashes  
In the windows operating environment - R will accept file names with forward 
slashes or back slashes but file names 
with back slashes must have double back slashes as the back slash is a special 
character in R. ie. 'C:/users/caroline' or 'C:\\\\users\\\\caroline'). 
The mac operating environment always uses a forward slash. 


## The Working Directory
R uses a default working directory. This is usually the users home folder if 
you are 
not using projects and the folder where the project file resides if you are in 
a project. The location of the working directory can be found with 
```{r}
getwd()
```
and set with
```{r, eval = F}
setwd('path_to_new_directory')
```

# R Basics

## Functions  
Function calls are the fundamental building blocks of R. A rigorous treatment of
function calls is beyond the scope of this course but for further reading, 
we recommend 
[Hadley Wickham's Advanced R: Functions](http://adv-r.had.co.nz/Functions.html). 
Everything that is done in R is done with a function. Functions take the form
function_name(function_arguments). Some functions have no arguments, for example
to get the current working directory (covered above):
```{r}
getwd()
```

Functions always have parenthesis, data objects do not.

## Data Objects  
Data objects are assigned with the assignment operator '<-' or '->'. The arrow points
to the object name. 
Attempting to call an object that has not been assigned will generate an error 
message. (User defined functions are also created using the assignment operator 
but that is beyond the scope of this course.) 

```{r}
Y <- 5
Y
27 -> Z
Z
X
```

Objects can  be assigned in succession. Assignment can also be done inside
expressions. 
```{r}
x <- y <- z <- 5
# assigns 5 to x, y, and z 
c(x, y, z)

x <- 2*(y <- 5)
# assigns 5 to y and 2*y = 10 to x
c(x, y)
```

## Inline Operators  
R contains the following built-in inline operators that can be used as is:
^, *, /, +, -, >, >=, <, <=, ==, !=, !, &, &&, |, ||, ~, <-, <<-, :, ::, :::, 
$, @.  
For example:
```{r}
1 + 3
2 * 6
x <- 4 + 8
x
3*4 == 6*2
13 <= 1
1 <= 13
5*2 == 10 & 3*4 <= 6
5*2 == 10 | 3*4 <= 6
```

## Installing Packages and Loading Packages
One of the features of R is the ability for users to create packages. Some 
packages come loaded with the base install of R. Others must be installed.
To install a package not included with base R, run 
install.package('package_name')
```{r, eval = F}
install.packages('ggplot2')
```

To load a previously installed package or a package that was installed with base
R for use in the current R session use library(package_name). 
For install.packages()
the package name must be in quotations, for library() either is fine.
```{r, eval = F}
library(ggplot2)
# or
library('ggplot2')
```

