---
title: "Intro to R/R Studio"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = TRUE)
knitr::opts_knit$set(root.dir = '../')
```

# Installing R and Rstudio
R can be installed from [here](https://cran.r-project.org). Download the base
version
for your operating system.  
RStudio can be installed from 
[here](https://www.rstudio.com/products/rstudio/download/#download). Download
the
installer for your platform.  

**R must be installed before installing RStudio**  

# R code and R output  
Throughout this course you will be presented with R code and R output. 
R code will be in boxes like this:
```{r, eval=FALSE}
print('This is some R code.')
```

This can be copied and pasted as is into your R script or R console. 

R output will be preceded by double hash marks and look like this:
```{r, echo=F}
print('This is R output.')
```

# R is case sensitive  
Function names, object names, etc. are all case senstive in R. This means that 
the function mean() $\neq$ Mean() $\neq$ MEAN() and data $\neq$ Data $\neq$ DATA.
```{r}
a <- 1
A <- 5
a
A
identical(a, A)
```

# File Paths  
A file path specifies the unique location of a file. For example the file path 
for this file is '/Users/caroline/Repositories/BIOS6605/Modules/Module_1.html'. 
R recognizes relative file paths. For example if the current working directory 
(see below) is '/Users/caroline/Repositories/BIOS6605/', R will allow you to 
access this file with the relative file path 'Modules/Module_1.html'.  

## Some Handy Shortcuts  
'~/' always represents a users home folder and '../' takes you back up one
folder level. '../' can be used in succession to go up multiple folder levels. 
For example this file's path can be represented as 
'~/Repositories/BIOS6605/Modules/Module_1.html' (where '/Users/caroline' is 
replaced with '~') '../' will take you to '/Users/caroline/Repositories/' (one 
level up from '/Users/caroline/Repositories/BIOS6605/') and '../../' will take you
to '/Users/caroline/', two levels up.

## Slashes  
In the windows operating environment R will accept file names with forward 
slashes or back slashes but file names 
with back slashes must have double back slashes as the back slash is a special 
character in R. ie. 'C:/users/caroline' or 'C:\\\\users\\\\caroline'). 
The mac operating environment always uses a forward slash. 


## The Working Directory
R uses a default working directory - usually the users home folder if you are 
not using projects and the folder where the project file resides if you are in 
a project. The location of the working directory can be found with 
```{r}
getwd()
```
and set with
```{r, eval = F}
setwd('path_to_new_directory')
```

# Functions  
Function calls are the fundamental building blocks of R. A rigorous treatment of
function calls is beyond the scope of this book but for further reading, 
we recommend 
[Hadley Wickham's Advanced R: Functions](http://adv-r.had.co.nz/Functions.html). 
Everything that is done in R is done with a function. Functions take the form
function_name(function_arguments). Some functions have no arguments, for example
to get the current working directory (covered above):
```{r}
getwd()
```

Functions always have parenthesis, objects do not.

# Objects  
Objects are assigned with '<-' or '->'. The arrow points to the object name. 
Attempting to call an object that has not been assigned will generate an error 
message. 

```{r}
Y <- 5
Y
27 <- Z
Z
X
```





