---
title: 'Module 1a: Entering Data'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = TRUE, warning  = TRUE)
knitr::opts_knit$set(root.dir = '../')
```


# Data Structures and Types
The most important types of data structures for our purposes are:  
* Vectors (1 dimensional)  
* Matrices (2 dimensional)  
* Arrays (n dimensional)  
* Data Frames  
* Lists  

The first 3 must be homogeneous and the last two may be heterogeneous. What does
this mean? There are four common types in R: logical, integer, double, and 
character. The homogeneous data structures must be all of one type. R will coerce
data into the most flexible type present if more than one type is used. Vectors 
are usually created with the function 'c()' (c for combine or concatenate). 

For example:
```{r}
# a vector with only logical types
logical_vect <- c(T, F, T, T)
logical_vect
typeof(logical_vect)
is.vector(logical_vect)

# a vector with only integer types
int_vect <- c(1L, 2L, 3L, 4L)
is.integer(int_vect)
typeof(int_vect)

# a mixed vector with logical and integers
mixed_vect1 <- c(T, 1L, 3L, F)
mixed_vect1
typeof(mixed_vect1)
```

Side-note: to specify an integer use the suffix capital L. 

```{r}
int_vect <- c(1L, 2L, 3L, 4L)
is.integer(int_vect)
typeof(int_vect)

double_vect <- c(1, 2, 3, 4)
is.integer(double_vect)
typeof(double_vect)

# a mixed vector with numbers and characters
mixed_vect2 <- c(0.1, 3, 'a')
mixed_vect2
typeof(mixed_vect2)
```

Lists and data frames allow for mixed data types. Lists can contain any 
type of data and lists with in lists do not need to be the same size. 
Lists are created with the list() function call. 
```{r}
list1 <- list('A', 'B', 3L, 3, T)
str(list1)

list2 <- list(T, F, 5L)
list_1and2 <- list(list1, list2)
str(list_1and2)
```
In a data frame, the 
individual variable can only contain one type - ie. character, integer, factor, 
list. However because data frames can contain lists - different types of data 
can be put in a list. All variables in a data frame must have the same length 
(the same number of observations). Note: because objects are converted to 
data frames with 1 variable before being combined if you want to add a list, 
it must be done separately. Data frames are created with the function 
data frame()

```{r}
df1 <- data.frame(V1 = c(1L, 2L, 3L), V2 = c(1, 0, 1), 
                  V3 = c(1L, 1, 2), V4 = c('Male', 'Female', 'Female'), 
                  V5 = list(list1, list2, list_1and2))
str(df1)
df2 <- data.frame(V1 = c(1L, 2L, 3L), V2 = c(1, 0, 1), 
                  V3 = c(1L, 1, 2), V4 = c('Male', 'Female', 'Female'))
df2$V5 <- list(list1, list2, list_1and2)
str(df2)
```
Character variables are automatically converted to factors unless 
stringsAsFactors is set to FALSE. (We get into factors below). 

```{r}
df3 <- data.frame(Gender = c('Male', 'Female', 'Female', 'Female'), 
                  Age = c(24, 24, 25, 24))
str(df3)

df4 <- data.frame(Gender = c('Male', 'Female', 'Female', 'Female'), 
                  Age = c(24, 24, 25, 24), stringsAsFactors = F)
str(df4)
```

Data frames can also be created by specifying object names. 
```{r}
Age <- c(25L, 30L, 56L)
Gender <- c('Male', 'Female', 'Male')
Weight <- c(160, 110, 220)
mydata1 <- data.frame(Age, Gender, Weight)
str(mydata1)
```
For this class, we will mostly be focused on data frames. 

## Factors
A factor is a special type of integer vector. Factor class vectors have a
special attribute: levels. It is likely that you will use factors extensively in
your analysis. They are ideal for categorical variables because a large number 
of statistical methods automatically recognize factors as categorical variables
and perform any necessary dummy coding behind the scenes. BIOS 6601 will cover 
cover the need for dummy coding in regression models later in the summer. 
For our purposes you need to know that factors are a special class of vector 
used for coding categorical variables. The levels attribute sets the allowed 
levels (classes/categories). Levels can be set when the factor is created or 
separately. The levels of a factor can be accessed with the levels function. 
Examples:

```{r}
x <- c('A', 'A', 'B', 'C')
x <- factor(x)
x
str(x)
class(x)
levels(x)
# You can add levels 
levels(x) <- c('A', 'B', 'C', 'D')
levels(x)
x
# note that setting the levels will completely replace the original levels
levels(x) <- c('Dog', 'Cat', 'Bird', 'Rabbit')
x # no information about the original levels/ values (A, B , and C) is retained
str(x)
# trying to replace the levels with fewer levels will give an error
# note this is the case even though x does not contain the 4th level
levels(x) <- c('A', 'B', 'C')

# trying to assign a value that is not in the levels will results in 
# an assignment of NA and will produce a warning
x[3] <- 'Monkey'
x
str(x)

# levels are ordered - if you change the order - you will change the level
# assigned to a data point. only the index of the level is preserved
x <- factor(c('A', 'A', 'B', 'C'))
x
levels(x) <- c('C', 'B', 'A')
x # notice this is now different from above
```
There are various methods for reordering levels and combining levels that we will
address in a later module. 

### The cut function  
Categorical variables can be easily created from continous variables using the 
cut function. The cut function accepts arguments for break points including 
the starting and ending points. To include all values below a certain value use 
'-Inf' as the starting point. To include all values above a certain value use 
'Inf' as the end point. By default categories are closed on the right (meaning 
it includes the lowest value but not the highest - for example 
cut(x, breaks = c(-Inf, 10, 20, Inf)) would generate intervals of less than or
equal to ten, greater than 10 to less than or equal to 20, and greater than 20. 
If the endpoints are not (-Inf, Inf), values outside of those endpoints will 
be categorigized as NA. Labels can be assigned using the labels option. You will
always have one less label than cutpoint. The resulting vector is a factor. 

```{r}
data("precip")
# Included dataset that has The average amount of precipitation (rainfall) in 
# inches for each of 70 United States (and Puerto Rico) cities.
# for more info see ?precip
precipcat <- cut(precip, breaks = c(0, 29, 37, 43, Inf), 
                 labels = c('Very Dry', 'Dry', 'Wet', 'Very Wet'))
table(precipcat, useNA = 'always')
precipcat <- cut(precip, breaks = c(0, 29, 37, 43, 65), 
                 labels = c('Very Dry', 'Dry', 'Wet', 'Very Wet'))
table(precipcat, useNA = 'always')
is.factor(precipcat)
```

### The table and prop.table functions
The table and prop.table function are very useful in examining categorical 
variables, making frequency tables and examining proportions. For right now, 
we will contstrain our discussion to one way and two-way frequencies. Later in 
the semester we will expand that to three-way frequecies and use these functions
to perform statisitcal tests. To obtain row, column or overall percentages use
prop.table with the table object. 

```{r}
# data frame with Gender and Smoking Status Randomly assigned
set.seed(123) # we will discuss setting seeds in a later module
Gender <- sample(c('Female', 'Male'), 50, replace = T)
set.seed(999)
Smoker <- sample(c('Smoker', 'NonSmoker'), 50, replace = T)
df5 <- data.frame(Gender, Smoker)
# look at Smoking Status by Gender
table(df5$Gender, df5$Smoker)
# there are 10 Female Non Smokers, 26 Non Smokers Overall and 
# 25 females overall
prop.table(table(df5$Gender, df5$Smoker))
# 20% of the overall population (50) are Female Non Smokers (10)
# note the sum of all percentages = 100

# margin = 1 for row percentages
prop.table(table(df5$Gender, df5$Smoker), margin = 1)
# 40% of the females overall (25) are non smokers(10)
# note the sum of both rows = 100

#margin  = 2 for column percentages
prop.table(table(df5$Gender, df5$Smoker), margin = 2)
# 38% of non-smokers overall (26) are female (10)
# note the sum of both columns = 100
```


## Logicals
Logicals have the special property that FASLE = 0 and TRUE = 1. This allows 
mathematical operations to be performed. 

```{r}
logical_vect <- c(T, F, T, T, F, T)
# get number of trues
sum(logical_vect)

# get percentage
sum(logical_vect)/length(logical_vect)
```

# Subsetting
To access individual elements of a data object use square brackets or the 
dollar sign (lists and data frames). Examples:
```{r}
# for one dimesional vectors
x <- c(1:10)
x
x[4]
x[5]
# you can not use $ outside of lists or data frames
x$5
```

```{r}
# for more than one use c()
x[c(1, 5, 7)]
# or sequences
x[1:6]

str(list1)
list1[1]
list1[1:3]

# for two dimensional objects - the first number is the row, the second
# is the column, to get all values in a row or a column use
# blanks
x <- matrix(c(1:10), ncol = 5)
x
x[1, 4]
x[, 4]
x[1, ]

df6 <- data.frame(x = c(1:10), z = letters[1:10], 
                  stringsAsFactors = F)
df6
df6[5:9, 1]
df6[8, 1:2]
# specifing indeices outside of the range will return an error
df6[2, 3:4]
```

```{r}
# lists and dataframes can also be subsetted using '$'
# for lists the list MUST be named (data frames are always named)
list1$`1`
names(list1) <- c('1', '2', '3', '4', '5')
list1
list1$`1`
df6$x
df6$z
```
For a more indepth discussion of subsetting we recommend 
[Advanced R: Subsetting](http://adv-r.had.co.nz/Subsetting.html)


