---
title: 'Module 3: Confidence Intervals'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = TRUE, warning  = TRUE)
knitr::opts_knit$set(root.dir = '../')
```
# Z scores (standard scores)
Recall that we can standardize values to a normal distribtion with mean = 0 and 
standard deviation = 1 by converting all of the values to z scores using the 
formula 
$$z = \frac{x - \mu}{\sigma}$$ 
where $\mu$ is the population mean and $\sigma$ is the population mean.  
In most 
cases the population mean and the population standard deviation are unknown and 
the sample mean and standard deviation are used. 
$$z = \frac{x - \bar{x}}{s}$$
R can quickly scale a variable for us using this formula.
```{r zscore}
age <- c(54, 45, 47, 40, 39, 53, 54, 51, 59, 71)
mean(age)
sd(age)
ageScaled <- (age - mean(age))/sd(age)
mean(ageScaled)
sd(ageScaled)
```

The new variable is approximately normally distributed with mean = 0 and 
standard deviation = 1. 

# Confidence Intervals
When reporting a parameter estimate, it is typical to include a measure of 
uncertainty with that point estimate. The preferred solution to this problem 
is to provide a confidence interval of a predetermined level of confidence, 
say (1-$\alpha$)%, which represents a range of values that has a specified 
probability of containing the real value of the parameter being estimated.  

You can use some simple R code to calculate Confidence Intervals (CI). To 
illustrate this, we are going to use some concepts we learned in Module 2 to 
generate a data set that involves 10 teachers and their annual salary. 

```{r teach}
Teacher <- 1:10
set.seed(123)
Salary <- rnorm(10, mean = 50500, sd = 18000)
# run mean and sd to see how close our actual mean and sd are 
# to those we specified
TeachData <- data.frame(Teacher, Salary)
TeachData
```

All of this code should be familiar to you. If not, please review Modules 1 & 2. 

Now lets find the confidence interval for the mean value of the teachers' salary. 
Recall that:
$$(1 - \alpha)\%CI = \bar{x}  \pm Z_{(1-\alpha/2)}\frac{\sigma}{\sqrt n} $$
To get $Z_{1-\alpha/2}$ we will use the qnorm() function. For the 95% CI, 
$\alpha = 0.05$. Therefore $1-\alpha/2 = 0.975$
```{r qnorm}
qnorm(.975)
# lets store this for ease of use
Z_a <- qnorm(.975)
```

Now we need to get $\bar{x}$ and $Z_{(1-\alpha/2)}\frac{\sigma}{\sqrt n}$. 
For this section we will use our
"known" population standard deviation ($\sigma$) of 18,000. 

```{r meansd}
xbar <- mean(TeachData$Salary)
SE <- Z_a * (18000/sqrt(10)) # where Z_a is from above and n = 10
```

Now to put it all together:
```{r ci}
LowerCI <- xbar - SE
LowerCI
UpperCI <- xbar + SE
UpperCI
```
Thus our 95% confidence interval is (
$`r format(LowerCI, big.mark = ',', digits = 7, nsmall = 2, trim = T)`, 
$`r format(UpperCI, big.mark  = ',', digits = 7, nsmall = 2, trim = T)`). 

## T distribution
Most of the time the population standard deviation is unknown and so we use the 
sample standard deviation. Then it is appropriate to use the t-distribution. 
The t-distibtion resembles the normal distribution but has fatter tails. The 
more degrees of freedom in a t-distribtion, the more closely it resembles the 
normal.

For more information on the r functions for the t-distribution:
```{r thelp}
?TDist
```

We will use the same formula as above with 2 small changes, we will replace
$Z_{1-\alpha/2}$ with $t_{(1-\frac{\alpha}{2}, n-1)}$ and $\sigma$ with $s$:
$$(1-\alpha)\%CI =\bar{x}\pm t_{(1-\frac{\alpha}{2}, n-1)}\frac{s}{\sqrt{n}} $$
Similar to before:

```{r qt}
qt(.975, 9) # qt function is the analog of qnorm for the t-distribtion
t_a <- qt(.975, 9)
```


```{r tsd}
xbar <- mean(TeachData$Salary)
SE_t <- t_a * (sd(TeachData$Salary)/sqrt(10))
```

Now to put it all together:
```{r tci}
LowerCI_t <- xbar - SE_t
LowerCI_t
UpperCI_t <- xbar + SE_t
UpperCI_t
```

Thus our 95% confidence interval is (
$`r format(LowerCI_t, big.mark = ',', digits = 7, nsmall = 2, trim = T)`, 
$`r format(UpperCI_t, big.mark  = ',', digits = 7, nsmall = 2, trim = T)`).

We can also use the t.test function to arrive at the same result. The t.test 
function performs the hypothesis test $H_0: \mu = 0$, but also generates a 
confidenve interval for $\mu$. 

```{r ttest}
t.test(TeachData$Salary, conf.level = 0.95)
```

Confirm this confidence interval is the same as above. 