---
title: 'Module 1: Entering Data'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = '../')
```


# Data Structures and Types
The most important types of data structures for our purposes are:  
* Vectors (1 dimensional)  
* Matrices (2 dimensional)  
* Arrays (n dimensional)  
* Data Frames  
* Lists  

The first 3 must be homogeneous and the last two may be heterogeneous. What does
this mean? There are four common types in R: logical, integer, double, and 
character. The homogeneous data structures must be all of one type. R will coerce
data into the most flexible type present if more than one type is used. Vectors 
are usually created with the function 'c()' (c for combine or concatenate). 

For example:
```{r}
# a vector with only logical types
logical_vect <- c(T, F, T, T)
logical_vect
typeof(logical_vect)
is.vector(logical_vect)

# a vector with only integer types
int_vect <- c(1L, 2L, 3L, 4L)
is.integer(int_vect)
typeof(int_vect)

# a mixed vector with logical and integers
mixed_vect1 <- c(T, 1L, 3L, F)
mixed_vect1
typeof(mixed_vect1)
```

Sidenote: to specify an integer use the suffix capital L. 

```{r}
int_vect <- c(1L, 2L, 3L, 4L)
is.integer(int_vect)
typeof(int_vect)

double_vect <- c(1, 2, 3, 4)
is.integer(double_vect)
typeof(double_vect)

# a mixed vector with numbers and characters
mixed_vect2 <- c(0.1, 3, 'a')
mixed_vect2
typeof(mixed_vect2)
```

For this class, we will mostly be focused on data frames. 



