---
title: 'Module 4: Data Exploration'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = TRUE, warning  = TRUE)
knitr::opts_knit$set(root.dir = '../')
```

# Data Exploration  
The primary task before analysis is exploring our data. This allows us to 
evaulate the appropriateness of statistical procedures, the linearity of 
relationships and explore how missing data may affect our results. In this 
module we will discover some of the tools available in R for data exploration.  

## Missingness  
Missing data points in R are represented by "NA". The function 'is.na()` is an 
extrememly useful way of idenitfiying missing values. is.na() will return TRUE 
for missing values and FALSE for non-missing values. 

```{r isna}
x <- c(2, 4, NA, 10, 11, NA, 3)
is.na(x)
# remember that we can select values where a function is equal to true using
# subsetting
x[is.na(x)]
x[!is.na(x)]
```

**A side note about NAs and subsetting**
Subsetting with NA will always return NA. For example:
```{r subsetna}
x
# here the equality NA < 5 returns NA so where NA is in x this returns NA
x[x < 5] 
```

```{r eval = F}
# what do you think this will return?
x < 5
# run it and find out - were you right?
```

Some functions functions will return NA if any NAs are present in their 
arguments. Examples include sum, mean, range, min, max, quantiles. 
This behaviour can be changed with the argument na.rm = T
```{r}
sum(x)
sum(x, na.rm = T)
range(x)
range(x, na.rm = T)
```

Read the help for any() and think about what you think the following code will 
return.
```{r eval = F}
any(x < 0)
any(x < 5)
```
Now run the code.  Was the result what you expected?  

We can also explore missingness in categorical variables. Add the useNA = 'ifany'
or useNA = 'always' argument to the table function to display missing values. 
```{r}
# to run these examples you will need to download the smokingdata from CANVAS
# and place it a folder titled data in your working directory
load('Data/SmokingData.rda')
table(SmokingData$Gender, SmokingData$Smoker)
table(SmokingData$Gender, SmokingData$Smoker, useNA = 'ifany')
table(SmokingData$Gender, SmokingData$Smoker, useNA = 'always')
```

What do you notice about the missing data?

We can also use addNA() and na.omit() to deal with missing data. By using addNA 
in our table function we can look at missing data for only one of our variables.
```{r}
table(SmokingData$Smoker, SmokingData$Education)
table(SmokingData$Smoker, SmokingData$Education, useNA = 'ifany')
table(SmokingData$Smoker, addNA(SmokingData$Education))
```

addNA() adds a level for missing data to a factor. 
```{r}
x <- addNA(SmokingData$Education)
str(x)
levels(x)
head(x, 20)
```

na.omit() allows us to select only complete cases. This will remove any 
observations is a data frame where any of the variables is missing. 
```{r}
summary(SmokingData)
completecases <- na.omit(SmokingData)
summary(completecases)
```

## Plots
R has a number of base plotting options and because we use code to create them, 
they are reproducible and change with the data!

```{r}
# cars is a built in R data set with speed and dist (it is included with the
# install - you don't need to do anything to use it) for more info:
?cars
```

First just a basic plot of speed v distance

```{r}
plot(cars)
```

A nice overview of the data, but let's spice it up a little. 
```{r}
plot(cars, 
     pch = 4, # sets the type of point - here will use a cross
     col = 'purple', # set the color - 
     # r will accept a wide variety of colour names as well as hexidecimal 
     # equivalents
     main = 'Speed vs Stopping Distance in the Cars data', # set the main title
     xlab = 'Speed (mph)', # set the x axis title
     ylab = 'Stopping Distance (ft)' # set the y axis title
     )
```
  
A complete list of available symbols](https://www.statmethods.net/advgraphs/parameters.html)  

Now, lets add a regression line.  

```{r}
plot(cars, 
     pch = 16, # let's the change the point to a filled circle
     col = '#0000cc', # set the color - 
     main = 'Speed vs Stopping Distance in the Cars data', # set the main title
     xlab = 'Speed (mph)', # set the x axis title
     ylab = 'Stopping Distance (ft)' # set the y axis title
     )
abline(lm(dist ~ speed, data = cars))
```

Now let's look at some other plots using the mtcars data set. 
```{r}
summary(mtcars)
# for details on this data set
?mtcars

# let's factor cyl, vs, and am. 
mtcars$cyl <- factor(mtcars$cyl, levels = c(4, 6, 8), 
                     labels = c('Four Cylinder', 
                                'Six Cylinder', 
                                'Eight Cylinder'))
mtcars$vs <- factor(mtcars$vs, levels = 0:1, 
                    labels = c('V-shaped', 'straight'))
mtcars$am <- factor(mtcars$am, levels = 0:1, 
                    labels = c('automatic', 'manual'))
summary(mtcars)
```

R will often guess the type of plot we want based on the data that was passed 
to it. For example in the plots above R guessed we wanted the first variable
in cars plotted against the second variable in cars and because they were both
continuous - it gave us a scatter plot. 

```{r}
plot(mtcars$cyl)
plot(mtcars$hp)
plot(mtcars$hp ~ mtcars$cyl)
plot(mtcars$cyl ~ mtcars$hp)
```

Does it appear there is a relationship between horse power and number of 
cylinders?

```{r}
plot(mtcars$am ~ mtcars$cyl)
plot(mtcars$hp ~ mtcars$wt)
```

Do you think there's an association between transmission type and number of 
cylinders? How about between horse power and weight?  

Let's look at a histogram of miles per gallon. 

```{r}
hist(mtcars$mpg, 
     col = '#33cc33', 
     main = 'Histogram of Miles per Gallon', 
     xlab = 'MPG')
```

Use ?hist, ?plot, and ?par to explore further graphing options. 
For further study check out [Quick-R: Graphs](https://www.statmethods.net/graphs/index.html)
and [Quick-R: Advanced Graphs](https://www.statmethods.net/advgraphs/index.html).  

### Saving Graphs
The easist way to save graphs is to use the export button on the plot viewer 
in Rstudio. 



