---
title: 'Module 2b: Random Allocation'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = TRUE, warning  = TRUE)
knitr::opts_knit$set(root.dir = '../')
```

# Random Selection
We can use random number functions in statistical packages to select a random 
sample from a population or to randomly allocate individuals to various 
interventions that we might want to compare. Random sampling or selection is an 
excellent way to reduce bias in an observational study. Random allocation in a 
clinical trial or comparison study is the optimal way to adjust for 
possible confounders that we cannot or do not want to measure for each 
individual in the study.  

**Example: Random Selection Of A Specified Number Of Records From A Larger 
Finite List**  

In this example, you are conducting a study looking at glucose measurements and
gender in 10,000 medical records. As part of the quality control checks 
for the study - the investigators
have asked you to randomly select 25 records so they can pull the original 
medical record and review that the numbers are accurate. 

```{r}
# for this example you will need to download the dummy glucose data set from
# canvas and place it in a folder named Data in your working directory
# we will then use the load function to load a previously saved r data object
load('Data/DummyGlucoseData.rda')
# this should've loaded a data frame called 
# gluc_data with 10,000 observations of 3 variables
head(gluc_data)
```

We will now use the sample function to select a sample of records. The sample 
function randomly samples from a finite list with equal probabilities of each
item being chosen with or without replacement. In this example, we do not want 
to sample with replacement because we do not want to select the same record 
more than once. This is equivalent to putting a ball with each record number
in a bag and then drawing out 25 without putting the ball back in the bag. 

```{r}
set.seed(222)
MRN_list <- sample(gluc_data$MRN, size = 25, replace = F)
# we now have a list of 25 medical record numbers for our investigators
# lets look at the first 5.
MRN_list[1:5]
```


```{r, eval = F}
# we can now export this list to a .csv file to give to the investigators
write.csv(MRN_list, file = 'Data/MRNsForInvestigators.csv')
```

We could also select a percentage of records by substituting size in the sample
function with an expression.
```{r}
set.seed(333)
MRN_list_pct <- sample(gluc_data$MRN, 0.20 * nrow(gluc_data), replace = F)
# the new function nrow() tells us how many rows are in a data frame
# we can multiply this by the desired percent, here = 20%
length(MRN_list_pct)
```

# Random Allocation
We will be exploring two ways of allocating subjects into different treatment 
groups (say A and B) for an intervention trial using:  
•	**Simple allocation** - the subjects are randomly assigned a course of treatment 
(either A or B) with no guarantee that they equally distributed among A and B  
or  
•	**Block Allocation or Stratified Random Sampling** - the subjects are assigned to 
group A or B with exactly half (50%) assigned to group A and half assigned to 
group B.


## Simple Random Allocation
Simple random allocation, or random assignment, is the method of randomly 
assigning subjects to treatment groups with indifference to how many in each 
group. Suppose you are comparing the effectiveness of two drugs, Drug A and 
Drug B, in treating gastro-esophageal reflux disorder (GERD). You decide to 
test these treatments on the next 100 GERD patients that walk into a clinic.  
You now have to decide in what order to give the two treatments so you don’t 
inadvertently create biases. The ‘runif()’ function in R can be used to 
randomly assign patients to one of two treatment groups, A and B provided we 
desire a simple random sample.

```{r}
# set up a data frame with subject ids
Simple <- data.frame(ID = 1:100)
# set seed for reproducibility
set.seed(333)
Simple$TrtGrp <- ifelse(runif(100) <= 0.5, 'A', 'B')
# lets take a look
head(Simple, 10)
table(Simple$TrtGrp)
# this list can now be given to the researchers for assignment
```

## Block Allocation and Stratified Random Sampling
Suppose instead we want to make sure _exactly_ half of our subjects are assigned
to treatment group A and half to treatment group B. Here we will use the sample
function (remember that with the sample function each item has an equal 
probability of being selected). 

```{r}
# set up a data frame with subject ids
# we will start with all subjects be set to treatment group 'B'. 
Block <- data.frame(ID = 1:100, TrtGrp = 'B', stringsAsFactors = F)
# We will then randomly select exactly half for treatment group 'A'. 
set.seed(4)
Block$TrtGrp[sample(1:100, size = 50, replace = F)] <- "A"
table(Block$TrtGrp)
head(Block, 10)
```

Suppose we want to make sure that each group has an equal number of males and
females. We can set up a data set with fifty males labeled one to fifty and
fifty females labeled one to fifty. 

```{r}
Stratified <- data.frame(ID = c(paste0('M', 1:50), paste0('F', 1:50)), 
                         Stratum = c(rep('Male', times = 50), 
                                     rep('Female', times = 50)),
                         TrtGrp = factor('B', levels = c('A', 'B')))
# randomly select 25 males for 'A'
set.seed(10)
Stratified$TrtGrp[sample(1:50, size = 25, replace = F)] <- 'A'
set.seed(15)
#randomly select 25 females for 'A'
Stratified$TrtGrp[sample(51:100, size = 25, replace = F)] <- 'A'
table(Stratified$TrtGrp, Stratified$Stratum)
head(Stratified)
tail(Stratified)
```

In the above code blocks - what is the purpose of stringsAsFactors = F and
factor('B', levels = c('A', 'B'))?
What to do you think would happen if you ran the following code instead?

```{r, eval = F}
# set up a data frame with subject ids
# we will start with all subjects be set to treatment group 'B'. 
Block <- data.frame(ID = 1:100, TrtGrp = 'B')
# We will then randomly select exactly half for treatment group 'A'. 
set.seed(4)
Block$TrtGrp[sample(1:100, size = 50, replace = F)] <- "A"
```

Run it and find out. Was it what you expected?

Can you figure out what the function paste0() does? If not try '?paste0'. 