---
title: 'Module 1b: Reading and Saving Data'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = TRUE, warning  = TRUE)
knitr::opts_knit$set(root.dir = '../')
```

# Intro 
Most of the time the data you wish to analyze will be in a format other than R. 
We will go over how to read these other file types into R. For additional 
file types, we recommend 
[Quick R: Importing Data](https://www.statmethods.net/input/importingdata.html).  

# read.table
read.table allows you to read in data that is in a table format from text files. 
```{r}
# this code to run you will need to download the effort.txt file from canvas
# and add it to a folder called data in your working directory
la_effort <- read.table('Data/effort.txt')

# you can also download data from a URL
la_data <- read.table("http://data.princeton.edu/wws509/datasets/effort.dat")
```

We will use the summary and head function to get some details about our new data
frame. 
```{r}
summary(la_effort)

# specifies the first 5 observations
head(la_effort, 5)

# let's rename the variables so they are more descriptive
names(la_effort) <- c('Country', 'Setting', 'Effort', 'Change')
head(la_effort, 3)

```

**Important Concepts:**   
  1.	Here R has put the country name as the first column. We could instead 
      specify that these be used as the row names with the option row.names = 1 
      (for the first column).   
  2.	You can always tell R explicitly whether or not you have a header by 
      specifying the optional argument header=TRUE or header=FALSE to the read.table 
      function.  
  3.	If your file does not have a header line, R will use the default variable
      names V1, V2, ..., etc. To override these default names use read.table's 
      optional argument col.names to assign variable names. This argument takes a 
      vector of names which can pre-exist or be built in the command.   


```{r}
# read in the same data as before but this time - set the 1st column to row
# names and set the column names when the data is read in
la_effort <- read.table('Data/effort.txt',
                        col.names = c('', 'Setting', 'Effort', 'Change'), 
                        row.names = 1)
head(la_effort, 6)
```

# read.csv
read.csv allows us to conveniently read in csv (comma separated values) files. 
Many applications produce data in .csv files.

```{r}
# as above you will need to download the relevant data (cancer_and_smoking.csv)
# file from canvas and add it to a folder called data in your working directory
smoking <- read.csv('Data/cancer_and_smoking.csv')
smoking
# notice the variable names - by default R assumes header = T which is not 
# the case here and so the first observation has been used to assign
# variable names
# we will fix this by exlicitly stating header = F
smoking <- read.csv('Data/cancer_and_smoking.csv', header = F)
smoking
# we can now set the variable names
names(smoking) <- c("State", "Cigarette", "Bladder", "Lung", "Kidney", 
                    "Leukemia")
smoking
```

# Merging Data sets
Two data sets can be merged with the merge function. The 'by' option determines
how the data is matched up. 'by.x' and 'by.y' can be used if the variable names
are not identical in both data sets. 'all.x', 'all.y' and 'all' can be used to 
specify if all observations from the first data frame, all observations from 
the second data frame or all observations should be used. 
```{r}
df1 <- data.frame(ID = c(2, 3, 1, 4, 5), 
                  hgt = c(70, 67, 64, 66, 72), 
                  wgt = c(180, 156, 128, 118, 202), 
                  race  = c('Cauc', 'Af.Am', 'Af.Am', 'Cauc', 'Asian'), 
                  year = c('Jr', 'Fr', 'Fr', 'Sr', 'So'), 
                  SAT = c(1080, 1210, 840, 1340, 880))
df1
df2 <-  data.frame(ID = c(1, 2, 3, 5, 6), 
                   sex = c('Male', "Male", "Female", "Male", "Female"), 
                   age = c(16, 15, 22, 14, 19))
df2

# default merge
df3 <- merge(df1, df2, by = 'ID')
df3

# include all observations
df4 <- merge(df1, df2, by = 'ID', all = T)
df4

# include all observations from df1
df5 <- merge(df1, df2, by = 'ID', all.x = T)
df5

# include all observations from df2
df6 <- merge(df1, df2, by = 'ID', all.y = T)
df6

# rename id in df2 so names don't match
names(df2) <- c('id', 'sex', 'age')
df7 <- merge(df1, df2, by.x = 'ID', by.y = 'id', all = T)
df7
```

# Ordering Data
Data can be ordered by a column or columns using the order function. 
```{r}
df1

# note that because we are ordering the rows - the order function goes
# where the row indecies would go and there is a ', ' to specify all 
# columns
df1ord <- df1[order(df1$SAT), ]
df1ord
# if the ordered object is not saved it will only be ordered printed in the 
# console - no permentant changes will be made ie.
df1[order(df1$SAT), ]
# displays the ordered data frame but does not save it

# the order function returns the indecies of the rows in order
order(df1$SAT)

# to order from higest to lowest use decreasing = T
df1[order(df1$SAT, decreasing = T), ]
```

# Saving Data
Individual data objects can be saved in files with the file extension .rda. 
Groups of data objects can be saved in files with the extension .RData.

```{r eval = F}
# to save an individual R data object, specify the object and the file name
save(la_effort, file = 'Data/la_effort.rda')

# to save a group of r objects, list them and then the file name
save(la_effot, la_data, file = 'Data/la_objs.RData')

# to save all objects in the environment
save.image('AllObjs.Rdata')
```

