---
title: 'Supplemental Module: ifelse'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = TRUE, warning  = TRUE)
knitr::opts_knit$set(root.dir = '../')
```

# ifelse  
First, take a look at the ifelse help. (I've also included it at the end of the module.) 
```{r}
?ifelse
```

**Key parts of this are:**

<tr valign="top"><td><code>yes</code></td>
<td>
<p>return values for true elements of <code>test</code>.</p>
</td></tr>
<tr valign="top"><td><code>no</code></td>
<td>
<p>return values for false elements of <code>test</code>.</p>
</td></tr>

and

<p>If <code>yes</code> or <code>no</code> are too short, their elements are recycled.
<code>yes</code> will be evaluated if and only if any element of <code>test</code>
is true, and analogously for <code>no</code>.
</p>

Now let's take a look at what happens when we use this for homework 2. We will 
shrink to just 20 observations for better understanding. 

```{r}
set.seed(33)
Gender <- ifelse(runif(20) < 0.6, 'Male', 'Female')
set.seed(33) # note the seed must be set EVERY time a random function is called
FemaleRMR <- rnorm(n = sum(Gender == 'Female'), mean = 1500, sd = 300)
set.seed(33)
MaleRMR <- rnorm(n = sum(Gender == 'Male'), mean = 2000, sd = 400)
set.seed(33)
RMR <- ifelse(Gender == 'Female', FemaleRMR, MaleRMR)
```

Now let's look at what happened:
```{r}
Gender
MaleRMR
RMR
```

What do you notice?
Let's look at the MaleRMR vector and the RMR vector only for males.
```{r}
MaleRMR
RMR[Gender == 'Male']
```

The RMR values were pulled only from MaleRMR where the indices of Gender == 'Male' were true. 
The position where Gender == 'Female' were skipped. The vector was then recycled because it was shorter than 
Gender. We can see the same thing with the females. 

```{r}
FemaleRMR
RMR[Gender == 'Female']
```

Another way to see the difference is to look at the means and standard deviations. 
```{r}
mean(MaleRMR)
mean(RMR[Gender == 'Male'])
mean(FemaleRMR)
mean(RMR[Gender == 'Female'])
sd(MaleRMR)
sd(RMR[Gender == 'Male'])
```

### To assign the vector FemaleRMR to females and MaleRMR to males we need subsetting. 


```{r}
RMR2 <- NULL # this just creates the object
# you can't subset an object until you've created it
# Alternatively you can use
RMR2 <- numeric(length = 20)
RMR2[Gender == 'Female'] <- FemaleRMR
RMR2[Gender == 'Male'] <- MaleRMR
```

Let's have a look:

```{r}
MaleRMR
RMR2[Gender == 'Male']
FemaleRMR
RMR2[Gender == 'Female']
mean(MaleRMR)
mean(RMR2[Gender == 'Male'])
mean(FemaleRMR)
mean(RMR2[Gender == 'Female'])
sd(MaleRMR)
sd(RMR2[Gender == 'Male'])
```


# ifelse help documentation from R:

<table width="100%" summary="page for ifelse {base}"><tr><td>ifelse {base}</td><td style="text-align: right;">R Documentation</td></tr></table>

<h2>Conditional Element Selection</h2>

<h3>Description</h3>

<p><code>ifelse</code> returns a value with the same shape as
<code>test</code> which is filled with elements selected
from either <code>yes</code> or <code>no</code>
depending on whether the element of <code>test</code>
is <code>TRUE</code> or <code>FALSE</code>.
</p>


<h3>Usage</h3>

<pre>
ifelse(test, yes, no)
</pre>


<h3>Arguments</h3>

<table summary="R argblock">
<tr valign="top"><td><code>test</code></td>
<td>
<p>an object which can be coerced to logical mode.</p>
</td></tr>
<tr valign="top"><td><code>yes</code></td>
<td>
<p>return values for true elements of <code>test</code>.</p>
</td></tr>
<tr valign="top"><td><code>no</code></td>
<td>
<p>return values for false elements of <code>test</code>.</p>
</td></tr>
</table>


<h3>Details</h3>

<p>If <code>yes</code> or <code>no</code> are too short, their elements are recycled.
<code>yes</code> will be evaluated if and only if any element of <code>test</code>
is true, and analogously for <code>no</code>.
</p>
<p>Missing values in <code>test</code> give missing values in the result.
</p>


<h3>Value</h3>

<p>A vector of the same length and attributes (including dimensions and
<code>"class"</code>) as <code>test</code> and data values from the values of
<code>yes</code> or <code>no</code>.  The mode of the answer will be coerced from
logical to accommodate first any values taken from <code>yes</code> and then
any values taken from <code>no</code>.
</p>


<h3>Warning</h3>

<p>The mode of the result may depend on the value of <code>test</code> (see the
examples), and the class attribute (see <code><a href="class.html">oldClass</a></code>) of the
result is taken from <code>test</code> and may be inappropriate for the
values selected from <code>yes</code> and <code>no</code>.
</p>
<p>Sometimes it is better to use a construction such as
</p>
<pre>  (tmp &lt;- yes; tmp[!test] &lt;- no[!test]; tmp)
</pre><p>, possibly extended to handle missing values in <code>test</code>.
</p>
<p>Further note that <code>if(test) yes else no</code>  is much more efficient
and often much preferable to <code>ifelse(test, yes, no)</code> whenever
<code>test</code> is a simple true/false result, i.e., when
<code>length(test) == 1</code>.
</p>
<p>The <code>srcref</code> attribute of functions is handled specially: if
<code>test</code> is a simple true result and <code>yes</code> evaluates to a function
with <code>srcref</code> attribute, <code>ifelse</code> returns <code>yes</code> including
its attribute (the same applies to a false <code>test</code> and <code>no</code>
argument).  This functionality is only for backwards compatibility, the
form <code>if(test) yes else no</code> should be used whenever <code>yes</code> and
<code>no</code> are functions.
</p>


<h3>References</h3>

<p>Becker, R. A., Chambers, J. M. and Wilks, A. R. (1988)
<em>The New S Language</em>.
Wadsworth &amp; Brooks/Cole.
</p>


<h3>See Also</h3>

<p><code><a href="Control.html">if</a></code>.
</p>


<h3>Examples</h3>

<pre>
x &lt;- c(6:-4)
sqrt(x)  #- gives warning
sqrt(ifelse(x &gt;= 0, x, NA))  # no warning

## Note: the following also gives the warning !
ifelse(x &gt;= 0, sqrt(x), NA)


## ifelse() strips attributes
## This is important when working with Dates and factors
x &lt;- seq(as.Date("2000-02-29"), as.Date("2004-10-04"), by = "1 month")
## has many "yyyy-mm-29", but a few "yyyy-03-01" in the non-leap years
y &lt;- ifelse(as.POSIXlt(x)$mday == 29, x, NA)
head(y) # not what you expected ... ==&gt; need restore the class attribute:
class(y) &lt;- class(x)
y
## This is a (not atypical) case where it is better *not* to use ifelse(),
## but rather the more efficient and still clear:
y2 &lt;- x
y2[as.POSIXlt(x)$mday != 29] &lt;- NA
## which gives the same as ifelse()+class() hack:
stopifnot(identical(y2, y))


## example of different return modes (and 'test' alone determining length):
yes &lt;- 1:3
no  &lt;- pi^(1:4)
utils::str( ifelse(NA,    yes, no) ) # logical, length 1
utils::str( ifelse(TRUE,  yes, no) ) # integer, length 1
utils::str( ifelse(FALSE, yes, no) ) # double,  length 1
</pre>

<hr /><div style="text-align: center;">[Package <em>base</em> version 3.5.0 <a href="00Index.html">Index</a>]</div>